---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# phonetisr: A Naive IPA Tokeniser

<!-- badges: start -->
`r badger::badge_devel(color = "orange")`
<!-- badges: end -->

This package is a (naive) tokeniser of phonetic transcriptions in the [International Phonetic Alphabet](https://www.internationalphoneticassociation.org/content/ipa-chart) (IPA).

With phonetisr, you can parse texts and word lists transcribed in IPA and tokenise them into phones so that you can perform quantitative analyses.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("stefanocoretta/phonetisr")
```

## Usage

```{r}
library(phonetisr)

# IPA strings to be tokenised
ipa <- c("pʰãkʰ", "tʰum̥", "ɛkʰɯ")

# List of character sequences to be considered single phones
ph <- c("pʰ", "tʰ", "kʰ", "ã", "m̥")

# Tokenise strings
phonetise(ipa, multi = ph)
```

## Roadmap

- [x] Scan for illegal (non-IPA) characters.
- [x] Provide a list of default multi-character phones.
- [ ] Functions for data import/export.
- [ ] Ignore diacritics.
